---
title: "Multi-temporal analysis of RS data via the 'ggridges' package"
author: "Alessio Monacchia"
date: "2024-01-22"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Visualizing vegetation phenological trend.

Here we use ridgeplots to display the variation of vegetation parameter (NDVI) related to phenology on an annual scale making use of the 'ggridges' R package.

We start by calling the libraries and importing the files:

```{r data, echo=TRUE}
## Libraries ####

library(raster)
library(terra)
library(purrr)
library(furrr)
library(ggridges)
library(tidyverse)

# Import data

setwd("/home/alessio/Tirocinio")

NDVI_300m <- list.files(pattern = "NDVI300_2014", recursive = T) %>%
  map(~rast(.))

# select one raster per each month to simplify calculations

NDVI_300m <- NDVI_300m[seq(1, length(NDVI_300m), by = 3)]
```

## Preprocessing

Here we crop the rasters to visualize the NDVI in Italy


```{r preprocessing, echo=TRUE}
## Pre-processing ####

# crop the rasters on Italian extent

Italy <- getData("GADM", country = "ITA", level = 0)

NDVI_300m_crop <- lapply(NDVI_300m, crop, Italy)

# plot the results

plot(NDVI_300m_crop[[1]])
```

Now, let's extract the data from the raster files and let's tidy it for the final ridgeplot!

```{r dataframe, echo=TRUE}
# Extract NDVI values in a list

values_NDVI_300m <- list()

for (i in 1:12) {
  
  new_object <- terra::as.data.frame(NDVI_300m_crop[[i]])
  
  values_NDVI_300m[[paste0("NDVI_300m_", i)]] <- new_object
  
  values_NDVI_300m[[i]]$month <- as.numeric(paste(i))  # create a column reporting the month of reference for each NDVI value
  
  rm(new_object)
  
}


# Merge list objects into a single dataframe

values_NDVI300_tot <- do.call(rbind, values_NDVI_300m) %>%
  
  subset(NDVI >= 0.2 | is.na(NDVI)) # set the condition excluding uninteresting cover types and NAs


# order rows according to month

values_NDVI300_tot <- values_NDVI300_tot[order(values_NDVI300_tot$month), ]


# check the varaible levels

unique(values_NDVI300_tot$month)


# turn variable month into a factor for the ridgeplot function requirements

values_NDVI300_tot$month <- as.factor(values_NDVI300_tot$month)


# Correct months names

values_NDVI300_tot$month <- fct_recode(values_NDVI300_tot$month, 
                                       January = "1", 
                                       February = "2",
                                       March = "3",
                                       April = "4",
                                       May = "5",
                                       June = "6",
                                       July = "7",
                                       August = "8",
                                       September ="9",
                                       October = "10",
                                       November = "11",
                                       December = "12")

```

## Ridgeplots

Now that the dataset is ready, let's go plotting our ridgeplots!

```{r ridgeplot, echo=TRUE}
## Ridgeplots ####

# free some memory

rm(NDVI_300m, NDVI_300m_crop, values_NDVI_300m)

gc()

# Simple NDVI ridgeplot with quantiles

ridgeplot_NDVI <- ggplot(values_NDVI300_tot, aes(NDVI, month, fill = factor(stat(quantile)))) +
  stat_density_ridges(
    geom = "density_ridges_gradient", calc_ecdf = TRUE,
    quantiles = 4, quantile_lines = TRUE)

print(ridgeplot_NDVI)

# Gradient ridgeplot type

ridgeplot_NDVI <- ggplot(values_NDVI300_tot, aes(x = `NDVI`, y = month, fill = after_stat(x))) +
  geom_density_ridges_gradient(scale = 3, rel_min_height = 0.01, gradient_lwd = 1.) +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_discrete(expand = expansion(mult = c(0.01, 0.25))) +
  scale_fill_viridis_c(name = "NDVI", option = "C") +
  labs(x = "NDVI",
       y = "Month",
    title = 'Phenological trend in Italy (2017)',
    subtitle = 'NDVI by month for 2017'
  ) +
  theme_ridges(font_size = 13, grid = TRUE)

print(ridgeplot_NDVI)
```
